Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-25 paper-new/concrete-relational-model.tex
move definition of the free composition of error domain relations out of appendix

Error domain relations are given by predomain relations that are
additionally a congruence with respect to the algebraic structures
$\theta$ and $\mho$.
\begin{definition}
  An \textbf{error domain relation} $d : B \rel B'$ is a monotone relation $d$ on the underlying predomains satisfying
  \begin{enumerate}
     \item ($\mho$ congruence): $\mho_B \binrel{d} \mho_{B'}$.
     \item ($\theta$ congruence): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t (\tilde{x}_t \binrel{d} \tilde{y}_t)$, then 
     $(\theta_B (\tilde{x}) \binrel{d} \theta_{B'} (\tilde{y}))$. 
  \end{enumerate}
\end{definition}
Again we have a reflexive error domain relation $r(B)$ given by the
ordering. The relational composition of two error domain relations in
general is not an error domain relation ($\theta$ congruence in
particular is not preserved), but we instead use an alternative ``free
composition'' $d \odot d'$ that is inductively generated from the
relational composition. Details of this construction are in the
appendix (Definition \ref{def:free-comp-ed-rel}).

We also will use the notion of a square $f \ltsq{c_i}{c_o} g$ between
predomain morphisms as well as squares $\phi \ltsq{d_i}{d_o} \psi$
between error domain morphisms, which are defined similarly, with no
extra interaction with the algebraic structure.

>>>>

Error domain relations are given by predomain relations that are
additionally a congruence with respect to the algebraic structures
$\theta$ and $\mho$.
\begin{definition}
  An \textbf{error domain relation} $d : B \rel B'$ is a monotone relation $d$ on the underlying predomains satisfying
  \begin{enumerate}
     \item ($\mho$ congruence): $\mho_B \binrel{d} \mho_{B'}$.
     \item ($\theta$ congruence): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t (\tilde{x}_t \binrel{d} \tilde{y}_t)$, then 
     $(\theta_B (\tilde{x}) \binrel{d} \theta_{B'} (\tilde{y}))$. 
  \end{enumerate}
\end{definition}
Again we have a reflexive error domain relation $r(B)$ given by the
ordering. The relational composition of two error domain relations in
general is not an error domain relation ($\theta$ congruence in
particular is not preserved), but we instead use an alternative ``free
composition'' $d \odot d' : B \rel B''$ that is inductively generated from the
relational composition. The rules are given in Figure \ref{fig:free-comp-ed-rel}.
% the appendix (Definition \ref{def:free-comp-ed-rel}).

\begin{figure}
  \begin{mathpar}
    \inferrule
    {x \mathbin{d} y \and y \mathbin{d'} z}
    {x \mathbin{(d \relcomp d')} z}

    \inferrule
    {x' \ltdyn_{B_1} x \and x \mathbin{(d \relcomp d')} z}
    {x' \mathbin{(d \relcomp d')} z}

    \inferrule
    {x \mathbin{(d \relcomp d')} z \and z \ltdyn_{B_3} z'}
    {x \mathbin{(d \relcomp d')} z'}

    \inferrule
    { }
    {\mho_{B_1} \mathbin{(d \relcomp d')} z}

    \inferrule
    {\later_t[ \tilde{x}_t \mathbin{(d \relcomp d')} \tilde{z}_t ] }
    {\theta_{B_1}(\tilde{x}) \mathbin{(d \relcomp d')} \theta_{B_3}(\tilde{z}) }
  \end{mathpar}
  \caption{The free composition of error domain relations.}
  \label{fig:free-comp-ed-rel}
\end{figure}

As in the previous section, we also have the notion of a square $f
\ltsq{c_i}{c_o} g$ between predomain morphisms, which is defined to mean that
for all $x \binrel{c_i} y$, we have $f(x) \binrel{c_o} g(y)$. Likewise, we have
squares $\phi \ltsq{d_i}{d_o} \psi$ between error domain morphisms, which are
defined similarly, with no extra interaction with the algebraic structure.
%
The free composition of error domain relations has the universal property that
$\phi \ltsq{d \relcomp d'}{d''} \phi'$ if and only if for any $x \binrel{d} y$
and $y \binrel{d'} z$, we have $\phi(x) \binrel{d''} \phi'(z)$.
